---
title: "Calibration model"
author: "Zoey Werbin"
date: "3/29/2019"
output: html_document
---

In this markdown document, we'll run through the creation of our models for the "historical" data from 5 NEON sites. We will model the ratio of bacteria and archaea (as one combined group) to fungi. Our current covariates are daily  precipitation and minimum daily temperature, both averaged to the month level.

We have other covariates that describe each site, such as mean canopy height, and climate descriptors (MAP and MAT). We also have covariates such as pH, C:N ratio, and soil temperature, that we hope to include if we receive more data from NEON - in this historical time-series, we only have 1-3 values for each of these soil chemical or physical characteristics. 

First, let's run the script that will download and aggregate our covariate data, and read in the output.

```{r setup}

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
print(getwd())
library(rjags)
library(daymetr)
library(zoo)
library("ecoforecastR")
library(dplyr)
```
```{r}
source("data_construction/aggregate_calibration_data.R")

### read in dataframe with response variable and covariates
df <- readRDS("data/calibration_model_data.rds")
```

Now, let's format our data object. First we'll look at the diagnostics for one of our sites, STER, in-depth. Later we'll loop through and look at the model fits for the other four sites.

```{r}
# extract data from STER site
site.data <- df[df$siteID=="STER",]
  y <- site.data$ratio
  
  # convert dateID to date
  time <- site.data$dateID
  time <- as.yearmon(time, format="%Y-%m")
  time <- as.Date(time, format="%Y-%m")
  
  # set up data object
  z <- cbind(rep(1,length(y)), site.data$min_temp.C_avg, site.data$precip.mm_avg)
  colnames(z) <- c("betaIntercept", "betaTmin", "betaPrecip")
  data <- list(y=log(y),n=length(y), x_ic = 0,tau_ic = 0.00001,a_obs=0.1,
                r_obs=0.1,a_add=0.1,r_add=0.1)
  data[["Z"]] <- z
```

Now let's set up our JAGS model code. We're including random effects for each sampling date
```{r}

ourmodel <- "model{

#### Priors
x[1] ~ dnorm(x_ic,tau_ic)
tau_obs ~ dgamma(a_obs,r_obs)
tau_add ~ dgamma(a_add,r_add)

#### Random Effects
tau_alpha~dgamma(0.1,0.1)
for(i in 1:n){
alpha[i]~dnorm(0,tau_alpha)
}

#### Fixed Effects
beta_IC~dnorm(0,0.001)
betaIntercept~dnorm(0,0.001)
betaTmin~dnorm(0,0.001)
betaPrecip~dnorm(0,0.001)

muTmin~dnorm(0,0.001)
muPrecip~dnorm(0,0.001)

tauTmin~dgamma(0.01,0.01)
tauPrecip~dgamma(0.01,0.01)

#### Data Model
for(t in 1:n){
y[t] ~ dnorm(x[t],tau_obs)
# missing data model:
Z[t,2] ~ dnorm(muTmin,tauTmin)
Z[t,3] ~ dnorm(muPrecip,tauPrecip)
}

#### Process Model
for(t in 2:n){
mu[t] <- beta_IC*x[t-1]  + betaIntercept*Z[t,1] + betaTmin*Z[t,2] + betaPrecip*Z[t,3] + alpha[t[i]]
x[t]~dnorm(mu[t],tau_add)
}

}"
```

Here we run the model and assess convergence, and check whether parameters are correlated.
```{r}
  j.model   <- jags.model (file = textConnection(ourmodel),
                           data = data,
                          # inits = init,
                           n.chains = 3)
  
  jags.out   <- coda.samples (model = j.model,
                              variable.names = c("tau_add","tau_obs", "beta_IC", "betaIntercept", "betaTmin",
                                                 "betaPrecip"),
                              n.iter = 10000)
  plot(jags.out)
  
  gelman.diag(jags.out)
  out <- as.matrix(jags.out)
  pairs(out)
```
All values are less than 1.05, so the model looks like it has converged. Now we'll run it again, monitoring the "x" variable this time. 
  
```{r}  
  jags.out   <- coda.samples (model = j.model,
                              variable.names = c("x","tau_add",
                              "tau_obs", "beta_IC", "betaIntercept", 
                              "betaTmin", "betaPrecip","tau_alpha","alpha"),
                              n.iter = 10000)
```
Let's visualize the model fit. This site has 10 dates, but other sites have even fewer, so this will likely be our best-fitting model.
```{r}

  # plot fit with confidence interval
  time.rng = c(1,length(time)) ## adjust to zoom in and out
  x.cols <- grep("^x",colnames(out)) ## grab all columns that start with the letter x
  ci <- apply(exp(out[,x.cols]),2,quantile,c(0.025,0.5,0.975)) ## model was fit on log scale
  
  plot(time, ci[2,], type='n', ylim=range(y,na.rm=TRUE), ylab="", 
       log='y', xlim=time[time.rng], xaxt="n", 
       main=paste0("Time-series for ",sites[s]))
  mtext(text = "Bacteria+Archaea : Fungi", side=2, line=2,cex=0.6)
  
  ## adjust x-axis labels
  axis.Date(1, at=seq(time[time.rng[1]],time[time.rng[2]],by='month'), format = "%Y-%m")
  ecoforecastR::ciEnvelope(time,ci[1,],ci[3,],col=ecoforecastR::col.alpha("lightBlue",0.75))
  points(time,y,pch="+",cex=0.5)

```


Now let's view all 5 sites together! We have to run this for more iterations, otherwise our HARV site (which only has 5 observations) won't converge.
```{r}

par(mfrow=c(5,1), mar=c(1.7,3.1,1.5,1.1), mgp=c(3,.6,0))
# loop through all 5 sites
for(s in 1:length(sites)) {

  site.data <- df[df$siteID==sites[s],]
  y <- site.data$ratio
  
  # convert dateID to date
  time <- site.data$dateID
  time <- as.yearmon(time, format="%Y-%m")
  time <- as.Date(time, format="%Y-%m")
  
   # set up data object
  z <- cbind(rep(1,length(y)), site.data$min_temp.C_avg, site.data$precip.mm_avg)
  colnames(z) <- c("betaIntercept", "betaTmin", "betaPrecip")
  data <- list(y=log(y),n=length(y), x_ic = 0,tau_ic = 0.00001,a_obs=0.1,
                r_obs=0.1,a_add=0.1,r_add=0.1)
  data[["Z"]] <- z
  
  
  j.model   <- jags.model (file = textConnection(ourmodel),
                           data = data,
                          # inits = init,
                           n.chains = 3)
  
  
  jags.out   <- coda.samples (model = j.model,
                              variable.names = c("x","tau_add",
                              "tau_obs", "beta_IC", "betaIntercept", 
                              "betaTmin", "betaPrecip",
                              "tau_alpha","alpha"),
                              n.iter = 80000)
  
  print(gelman.diag(jags.out))
  
  # plot fit with confidence interval
  time.rng = c(1,length(time)) ## adjust to zoom in and out
  out <- as.matrix(jags.out)
  x.cols <- grep("^x",colnames(out)) ## grab all columns that start with the letter x
  ci <- apply(exp(out[,x.cols]),2,quantile,c(0.025,0.5,0.975)) ## model was fit on log scale
  
  plot(time, ci[2,], type='n', ylim=range(y,na.rm=TRUE), ylab="", 
       log='y', xlim=time[time.rng], xaxt="n", 
       main=paste0("Time-series for ",sites[s]))
  mtext(text = "Bacteria+Archaea : Fungi", side=2, line=2,cex=0.6)
  
  ## adjust x-axis labels
  axis.Date(1, at=seq(time[time.rng[1]],time[time.rng[2]],by='month'), format = "%Y-%m")
  ecoforecastR::ciEnvelope(time,ci[1,],ci[3,],col=ecoforecastR::col.alpha("lightBlue",0.75))
  points(time,y,pch="+",cex=0.5)
}
```
