---
title: "Forecasting model"
author: "Zoey Werbin, Ryan Quinn, Steve Gougherty, Yetianjian Wang"
date: "3/29/2019"
output: html_document
---

In this markdown document, we'll run through a forecast using our dynamic linear models for the "historical" data from one NEON site, STER. We will model the ratio of bacteria and archaea (as one combined group) to fungi. Our current covariates are daily precipitation and minimum daily temperature, both averaged to the month level.

```{r setup}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
#setwd("..")
print(getwd())
```
Load directories:
```{r, echo=FALSE}
library(rjags)
library(daymetr)
library(zoo)
library(ecoforecastR)
library(dplyr)
```

Now, let's load our data object.

```{r}
# source file that downloads/aggregates all of the data - only need to run once. takes less than 10 min on a mac, takes much longer on non-Macs due to something (?) with the neonUtilities pacakge.
if(!file.exists("data/cal.jags.out.rds")){
  cat("Let's run the calibration model script before creating these forecasts.")
    source("analysis/calibration_model.R")
}
### read in dataframe with response variable and covariates
stopifnot(file.exists("data/cal.jags.out.rds"))
jags.out <- readRDS("data/cal.jags.out.rds")
```

Let's look at our JAGS model code. We're including random effects for each sampling date, and fixed effects of minimum temperature and precipitation.

```{r}
ourmodel <- "  model{
  
#### Priors
x[1] ~ dnorm(x_ic,tau_ic)
tau_obs ~ dgamma(a_obs,r_obs)
tau_add ~ dgamma(a_add,r_add)

#### Random Effects
  tau_alpha~dgamma(0.1,0.1)
  for(i in 1:n){                  
  alpha[i]~dnorm(0,tau_alpha)
  }

#### Fixed Effects
beta_IC~dnorm(0,0.001)
betaIntercept~dnorm(0,0.001)
betaTmin~dnorm(0,0.001)
betaPrecip~dgamma(0.01,0.01)
muTmin~dnorm(0,0.001)
muPrecip~dnorm(0,0.001)
tauTmin~dgamma(0.01,0.01)
tauPrecip~dgamma(0.01,0.01)


#### Data Model
for(t in 1:n){
OBS[t] ~ dnorm(x[t],tau_obs)
Z[t,2] ~ dnorm(muTmin,tauTmin)
Z[t,3] ~ dnorm(muPrecip,tauPrecip)
}

#### Process Model
for(t in 2:n){
mu[t] <- beta_IC*x[t-1]  + betaIntercept*Z[t,1] + betaTmin*Z[t,2] + betaPrecip*Z[t,3] + alpha[t[i]]
x[t]~dnorm(mu[t],tau_add)
}

}"
```


Let's take a look at our driver ensembles for our calibration period, sourced from the ERA5 ensembles (n=10):
```{r}
ppt <- readRDS("data/climate_ensembles.rds") # created using load_ERA5 function - only runs on SCC
# view temperature ensemble
temp <- ppt[[1]]
plot(1:length(temp[1,]),temp[1,],type='n',ylim=range(temp[1,]),xlab="time",ylab="temperature (c)")
for(i in 1:10){
  lines(1:length(temp[1,]),temp[i,],lwd=0.5,col="grey")
}

#view precipitation ensemble
precip <- ppt[[2]]
plot(1:length(precip[1,]),precip[1,],type='n',ylim=range(precip[1,]),xlab="time",ylab="precip (cm)")
for(i in 1:10){
  lines(1:length(precip[1,]),precip[i,],lwd=0.5,col="grey")
}
```

Load in our calibration model data to plot our real points:
```{r}
real_data <- readRDS("data/calibration_model_data.rds")
STER <- real_data[real_data$siteID=="STER",]

mc3.out <- jags.out
out = list(params=NULL,predict=NULL,model=ourmodel,data=data)
 mfit = as.matrix(mc3.out,chains=TRUE)
 pred.cols = union(grep("x[",colnames(mfit),fixed=TRUE),grep("mu[",colnames(mfit),fixed=TRUE))
 chain.col = which(colnames(mfit)=="CHAIN")
 out$predict = mat2mcmc.list(mfit[,c(chain.col,pred.cols)])
 out$params   = mat2mcmc.list(mfit[,-pred.cols])


# settings
NT = 10
s <- 1             ## Focal site for forward simulation
Nmc = 1000         ## set number of Monte Carlo draws
ylim = c(exp(4),exp(30))  ## set Y range on plot
N.cols <- c("black","red","green","blue","orange") ## set colors
trans <- 0.8       ## set transparancy
time = 1:(NT*2)    ## total time
time1 = 1:NT       ## calibration period
time2 = time1+NT   ## forecast period
NS <- 10

ppt_ensemble <- readRDS("data/climate_ensembles.rds")
ppt_ensemble <- ppt_ensemble[[2]]
# 
ppt.mean <- matrix(apply(ppt_ensemble,2,mean),1,NT) ## driver
## parameters
params <- as.matrix(out$params)
param.mean <- apply(params,2,mean)
## initial conditions
IC.orig <- as.matrix(out$predict)
NT <- 10
NS <- 10
s <- 5
ci <- apply(as.matrix(out$predict),2,quantile,c(0.025,0.5,0.975))

```

Here's the function we'll use for our forecast, analogous to the process model in our JAGS code. Right now we just have the precipitation driver included for the sake of developing the uncertainty analysis code - we will add Tmin back into this script later.
```{r}
# our function
forecastN <- function(IC,betaIntercept,alpha,betaPrecip,beta_IC,ppt,Q=0,n=Nmc) {
  N <- matrix(NA,n,NT)  ## storage
  Nprev <- IC           ## initialize
  N[,1] <- Nprev
  for(t in 2:NT){
      mu = beta_IC*Nprev + betaIntercept*1 + betaPrecip*ppt[,t] + alpha  
      N[,t] <- rlnorm(n,log(mu),Q)  
      Nprev <- N[,t]        ## update IC
  }
  return(N)
}

plot.run <- function(ylim=c(4,31)){
  sel = seq(1,10)
  plot(time,time,type='n',ylim=ylim, ylab="N")
  ecoforecastR::ciEnvelope(time1,ci[1,sel],ci[3,sel],col=col.alpha("lightBlue",0.6))
  lines(time1,ci[2,sel],col="blue")
  points(time1,log(STER[,"ratio"]))
}
```


We start with a deterministic forecast (every input is a mean, rather than full distribution):
```{r}
ci <- apply(as.matrix(out$predict),2,quantile,c(0.025,0.5,0.975))


N.det <- forecastN(IC=mean(IC.orig[,"x[10]"]),
                   betaIntercept=param.mean["betaIntercept"],
                   alpha=param.mean["tau_alpha"],
                   betaPrecip=param.mean["betaPrecip"],
                   beta_IC = param.mean["beta_IC"],
                   ppt=ppt.mean,
                   Q=0,  ## process error off
                   n=1)

## Plot run
plot.run()
lines(time2,N.det,col="purple",lwd=3)
```

Here we vary our initial conditions but hold everything else constant:
```{r}
prow = sample.int(nrow(params),Nmc,replace=TRUE)

N.I <- forecastN(IC=IC.orig[prow,"x[10]"],  ## sample IC
                   betaIntercept=param.mean["betaIntercept"],
                   alpha=param.mean["tau_alpha"],
                   betaPrecip=param.mean["betaPrecip"],
                   beta_IC = param.mean["beta_IC"],
                   ppt=ppt.mean,
                   Q=0,
                   n=Nmc)

# create plot
plot.run()
N.I.ci = apply(N.I,2,quantile,c(0.025,0.5,0.975))
ecoforecastR::ciEnvelope(time2,N.I.ci[1,],N.I.ci[3,],col=col.alpha(N.cols[1],trans))
lines(time2,N.I.ci[2,],lwd=0.5)
```


Here we provide the full distribution of parameters, to assess parameter uncertainty:
```{r}
N.IP <- forecastN(IC=IC.orig[prow,"x[10]"],  ## sample IC
                  betaIntercept=params[prow,"betaIntercept"],
                  alpha=params[prow,"tau_alpha"],
                   betaPrecip=params[prow,"betaPrecip"],
                   beta_IC = params[prow,"beta_IC"],
                   ppt=ppt.mean,
                   Q=0,
                   n=Nmc)

## Plot run
plot.run()
N.IP.ci = apply(N.IP,2,quantile,c(0.025,0.5,0.975), na.rm=T)
ecoforecastR::ciEnvelope(time2,N.IP.ci[1,],N.IP.ci[3,],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,N.I.ci[1,],N.I.ci[3,],col=col.alpha(N.cols[1],trans))
lines(time2,N.I.ci[2,],lwd=0.5)
```

Now for driver uncertainty (precipitation, which came from the 10 ERA5 ensembles):
```{r}
## sample driver rows
drow = sample.int(nrow(ppt_ensemble),Nmc,replace=TRUE)

N.IPD <- forecastN(IC=IC.orig[prow,"x[10]"],  ## sample IC
                  betaIntercept=params[prow,"betaIntercept"],
                  alpha=params[prow,"tau_alpha"],
                   betaPrecip=params[prow,"betaPrecip"],
                   beta_IC = params[prow,"beta_IC"],
                   ppt=ppt_ensemble[drow,],   ## Sample drivers
                   Q=0,
                   n=Nmc)

## Plot run
plot.run()
N.IPD.ci = apply(N.IPD,2,quantile,c(0.025,0.5,0.975), na.rm=TRUE)
ecoforecastR::ciEnvelope(time2,N.IPD.ci[1,],N.IPD.ci[3,],col=col.alpha(N.cols[3],trans))
ecoforecastR::ciEnvelope(time2,N.IP.ci[1,],N.IP.ci[3,],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,N.I.ci[1,],N.I.ci[3,],col=col.alpha(N.cols[1],trans))
lines(time2,N.I.ci[2,],lwd=0.5)
```
Our driver uncertainty is very small - negligible compared to the other uncertainties, possibly because driver values were well-constrained, i.e. ensembles mostly agreed with each other.
 
Process error uncertainty:
```{r}
## process error samples
Qmc <- 1/sqrt(params[prow,"tau_add"])  ## convert from precision to standard deviation

N.IPDE <- forecastN(IC=IC.orig[prow,"x[10]"],  ## sample IC
                  betaIntercept=params[prow,"betaIntercept"],
                  alpha=params[prow,"tau_alpha"],
                   betaPrecip=params[prow,"betaPrecip"],
                   beta_IC = params[prow,"beta_IC"],
                   ppt=ppt_ensemble[drow,],   ## Sample drivers
                   Q=Qmc,
                   n=Nmc)

## Plot run
plot.run()
N.IPDE.ci = apply(N.IPDE,2,quantile,c(0.025,0.5,0.975), na.rm=TRUE)
ecoforecastR::ciEnvelope(time2,N.IPDE.ci[1,],N.IPDE.ci[3,],col=col.alpha(N.cols[4],trans))
ecoforecastR::ciEnvelope(time2,N.IPD.ci[1,],N.IPD.ci[3,],col=col.alpha(N.cols[3],trans))
ecoforecastR::ciEnvelope(time2,N.IP.ci[1,],N.IP.ci[3,],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,N.I.ci[1,],N.I.ci[3,],col=col.alpha(N.cols[1],trans))
lines(time2,N.I.ci[2,],lwd=0.5)
```

This plot is misleading. Let's zoom out a bit...
```{r}
plot.run(ylim=c(-5,1000))
N.IPDE.ci = apply(N.IPDE,2,quantile,c(0.025,0.5,0.975), na.rm=TRUE)
ecoforecastR::ciEnvelope(time2,N.IPDE.ci[1,],N.IPDE.ci[3,],col=col.alpha(N.cols[4],trans))
ecoforecastR::ciEnvelope(time2,N.IPD.ci[1,],N.IPD.ci[3,],col=col.alpha(N.cols[3],trans))
ecoforecastR::ciEnvelope(time2,N.IP.ci[1,],N.IP.ci[3,],col=col.alpha(N.cols[2],trans))
ecoforecastR::ciEnvelope(time2,N.I.ci[1,],N.I.ci[3,],col=col.alpha(N.cols[1],trans))
lines(time2,N.I.ci[2,],lwd=0.5)
```

Something's wrong. But we don't know what it is, so let's partition the variance anyways!

We don't have a random site effect because our model isn't hierarchical. We're just looking at one NEON site (STER) right now. 
```{r}
### calculation of variances
varI     <- apply(N.I,2,var, na.rm=T)
varIP    <- apply(N.IP,2,var, na.rm=T)
varIPD   <- apply(N.IPD,2,var, na.rm=T)
varIPDE  <- apply(N.IPDE,2,var, na.rm=T)
#varIPDEA <- apply(N.IPDEA,2,var) # we only have one site
varMat   <- rbind(varI,varIP,varIPD,varIPDE)#,varIPDEA)

## in-sample stacked area plot
V.pred.rel.in <- apply(varMat,2,function(x) {x/max(x)})
plot(time2,V.pred.rel.in[1,],ylim=c(0,1),type='n',main="Relative Variance: In-Sample",ylab="Proportion of Variance",xlab="time")
ciEnvelope(time2,rep(0,ncol(V.pred.rel.in)),V.pred.rel.in[1,],col=N.cols[1]) # IC error = black
ciEnvelope(time2,V.pred.rel.in[1,],V.pred.rel.in[2,],col=N.cols[2]) # parameter error = red
ciEnvelope(time2,V.pred.rel.in[2,],V.pred.rel.in[3,],col=N.cols[3]) # driver error = green
ciEnvelope(time2,V.pred.rel.in[3,],V.pred.rel.in[4,],col=N.cols[4]) # process error = blue
legend("topright",legend=c("Process","Driver","Parameter","InitCond"),col=rev(N.cols[-5]),lty=1,lwd=5)
```

We have almost no driver uncertainty, so it was too small to plot. Parameter uncertainty is large, but it is still dwarfed by process error. 

